---
title: "Top 10 Stocks Analysis"
author: "Simon Macharia"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(lubridate)
library(ggplot2)
library(zoo)
library(reshape2)
library(corrplot)
library(readr)
library(scales)
```

# 1. ASK - Business Task

## 1.1. Key Questions:

-   What stock has performed best over time.
-   What are the volatility patterns of each stock.
-   Which stocks are most correlated, suggesting diversification opportunities.
-   Are there trends or seasonality in returns or volume traded.
-   How can we identify overvalued or undervalued stocks.

## 1.2. Business Goal:

To assess performance, trends, volatility, and correlation among 10 different stocks to inform investment or portfolio decisions.

# 2. PLAN - Data Requirements

The data to be used will include top 10 stocks in the last five years.\
That data is recorded in daily time frame trend.\

## Key performance indicators.

-   Daily return.
-   Cumulative return.
-   Volatility (standard deviation).
-   Sharpe ratio.
-   Volume trends.

**Data**: `10_stocks_data.csv`

# 3. PROCESS - Clean and Prepare the data

## 3.1 Load data

```{r load-data}
stocks <- read_csv("10_stocks_data.csv")
```

## 3.2 Daily Returns

```{r daily-returns}
stocks <- stocks %>% 
  group_by(symbol) %>% 
  mutate(
    previous_day_adjusted = lag(adjusted),
    daily_return = round((adjusted/ lag(adjusted) - 1) * 100, 2)
    ) %>% 
  arrange(symbol, date)

stocks
```

# 4. ANALYZE - Explore the Data

## 4.4. Time series analysis

### 4.4.1. Monthly trends

```{r daily-trend}
monthly_trend <- stocks %>%
  mutate(
    year = year(date),
    month = month(date, label = TRUE)
  ) %>%
  group_by(symbol,year,month) %>%
  arrange(date) %>%
  summarise(
    monthly_open = first(open),                   
    monthly_close = last(close)
  ) %>% 
  mutate(
    monthly_returns = round((monthly_close/lag(monthly_close) -1)* 100, 2)
  ) %>% 
  select(symbol, year, month, monthly_open, monthly_close, monthly_returns)

monthly_trend
```

### 4.4.2. Yearly trends

```{r yearly-trend}
yearly_trend <- stocks %>% 
  mutate(
    year = year(date)
  ) %>% 
  group_by(symbol, year) %>% 
  summarise(
    yearly_open = first(open),
    yearly_close = last(close)
  ) %>% 
  mutate(
    previous_close = lag(yearly_close),
    yearly_returns = round((yearly_close/lag(yearly_close) - 1)* 100, 2)
  ) %>% 
  select(symbol, year, yearly_open, yearly_close, previous_close, yearly_returns)

yearly_trend
```

## 4.5. Volatility (Rolling Std Dev)

```{r month-volatility}
yearly_volatility <- stocks %>% 
  mutate(
    year = year(date)
  ) %>% 
  group_by(year, symbol) %>% 
  mutate(rolling_volatility = rollapply(daily_return, width = 30, FUN = sd, fill = NA)) %>% 
  summarise(
    avg_volatility = mean(rolling_volatility, na.rm = TRUE),
    max_volatility = max(rolling_volatility, na.rm = TRUE),
    .groups = "drop"
  ) %>% 
  arrange(year, desc(avg_volatility))

yearly_volatility
```

## 4.6 Volume Analysis

```{r volume}
volume <- stocks %>% 
  mutate(
    year = year(date)
  ) %>% 
  group_by(year, symbol) %>% 
  summarise(
    avg_volume = mean(volume, na.rm = TRUE),
    max_volume= max(volume, na.rm = TRUE),
    .groups = "drop"
  ) %>% 
  arrange(year, desc(avg_volume))
  
volume
```

## 4.7 Correlation Analysis

```{r correlation}
correlation_data <- stocks %>% 
  mutate(
    daily_range = high - low,
    roll_volatility = rollapply(daily_return, width = 30, FUN = sd, fill = NA, align = "right"),
    roll_volume = rollapply(volume, width = 30, FUN = mean, fill = NA, align = "right")
  ) %>% 
  select(symbol, close, daily_return, daily_range, volume, roll_volume, roll_volatility ) %>% 
  drop_na()

cor_matrix <- correlation_data %>%
  ungroup() %>% 
  select(where(is.numeric)) %>%
  cor(use = "complete.obs")

cor_matrix

melted_mat <- melt(cor_matrix)

melted_mat

```

# 5. SHARE - Visualizations

## 5.1. Cumulative performance over time

**Chart**: Line chart for cumulative returns over time for each stock

```{r returns-overtime}
stocks %>%
  group_by(symbol) %>%
  filter(symbol %in% c('TSLA', 'META', 'NVDA')) %>% 
  arrange(date) %>%
  mutate(
    daily_return_clean = ifelse(is.na(daily_return), 0, daily_return),
    cumulative_return = cumsum(daily_return_clean)
  ) %>%
  ggplot(aes(x = date, y = cumulative_return, color = symbol)) +
  geom_line() +
  scale_x_date(date_breaks = "6 months", date_labels = "%b %Y") + 
  labs(
    title = "Cumulative Returns Over Time",
    subtitle = "Top 10 Stocks | Adjusted for Daily Returns",
    y = "Cumulative Return (%)",
    x = "Date",
    color = "Stock Symbol"
  ) +
  theme_minimal(base_size = 10) + 
  theme(
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(size = 12, color = "gray40"),
    axis.text.x = element_text(angle = 45, hjust = 1),  # rotated x labels
    legend.position = "right",
    legend.title = element_text(face = "bold")
  )
```

## 5.2. Volatility line plot

**Chart**:Line plot of rolling volatility (30-day) per stock

```{r volatility-plot}
stocks %>%
  mutate(year = year(date)) %>%
  filter(symbol %in% c('TSLA', 'META', 'NVDA')) %>%
  mutate(rolling_volatility = rollapply(daily_return, width = 30, FUN = sd, fill = NA)) %>%
  ggplot(aes(x = date, y = rolling_volatility, colour = symbol)) +
  geom_line(alpha = 0.8, linewidth = 0.8) +
  scale_x_date(date_breaks = "6 months", date_labels = "%b %Y") +
  labs(title = "Rolling 30-Day Volatility", y = "Volatility (SD)", x = "Date") +
  theme_minimal()+
  theme(
    plot.title = element_text(face = "bold", size = 16),
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "right"
  )
```

**Chart**: Bar plot for average volatility

```{r avg-volatility-plot}
monthly_volatility %>% 
  ggplot(aes(x = factor(year), y = avg_volatility, fill = symbol))+
  geom_col(position = "dodge")+
  labs(title = "Avarage volatility by year", x = "Year",y = "Average Volatility")+
  theme_minimal()
```

## 5.3. Yearly returns plot

**Chart**:Bar plot of yearly returns, faceted by stock.

```{r retrns-plot}
yearly_trend %>% 
  ggplot(aes(x = factor(year),y = yearly_returns, fill = symbol))+
  geom_bar(stat = "identity", position = position_dodge(width = 0.8),width = 1.0)+
  labs(title = "Yearly Returns by Stock", x = "Year", y = "Returns (%)")+
  theme_minimal()+
  scale_x_discrete(expand = c(0,0))
```

## 5.4. Correlation plot

```{r cor-plot}
corrplot(cor_matrix, method = "color", type = "upper", tl.cex = 0.8, tl.col = "black", addCoef.col = "black")

```

## 5.5. AVG Volume trend

**Chart**: Bar plot of average volume per year to show trading interests and\
liquidity patterns.

```{r volume-plot}
volume %>% 
  ggplot(aes(x = factor(year), y = avg_volume, fill = symbol))+
  geom_col(position = "dodge")+
  scale_y_continuous(labels = label_number(scale = 1/1e+06, suffix = "M"))+
  labs(title = "Average trading volume by year", x = "Year", y = "Average volume in(millions)")+
  theme_minimal()
```
